---
title: "Introduction to R"
output: html_document
---

# R-markdown

Here I will write normal text and instructions in *markdown*
language.
When there's a chunk of R code, it will be surrounded
by the symbols ```{r} and ```.

To run a line of R code, go to that line and hit Control+Enter.
The result will show up in the Rstudio Console and also below the 
R code chunk.

# R basics

## R as a calculator

```{r}
1+1
10*2
1+1*10
# R does math. By the way, lines starting with "#" are comments.
# This means that R will just ignore them.
3/4 # I can also add comments on the same line as code.
# Everything after "#" on that line will be ignored
```

## variables

### assignment

We can also store values in variables, which in R are called objects.
We call that assignment. 
The assignment symbol in R is `<-`.

```{r}
x <- 10
y <- 2+1
z <- x*y
z # if I just type a variable, R prints out its value.
```

### other types of objects: strings, boolean, vectors, matrices, dataframes

- Strings are aways in quotations

```{r}
name <- "Debora" # strings must always be surrounded by quotes
#Debora # if not, R thinks this is an object name and will complain if it doesn't exist
```

- Booleans are objects that take two values: `TRUE` or `FALSE`
  - they are the result of boolean operations such as `>`, `<` and `==`

```{r}
1==1
2>3
```
  - We can combine boolean values with the `&` (and) and `|` (or) operators
```{r}
bool1 <- 1==1 #this is TRUE
bool2 <- 2>3 #this is FALSE

bool1and2 <- bool1 & bool2 # are both true? FALSE
bool1and2

bool1or2 <- bool1 | bool2 # is either one true? TRUE
bool1or2
```
  
- Vectors are a sequence of objects
    - all objects must be of the same type
    - you can create them with the concatenation function, `c()`

```{r}
vec <- c(3,5,12,1345,x,y)
vec
# what happens if you add a number with decimal places?
# what if you add a string?
```
    - you can also create a vector that is a sequence of numbers
      with the colon operator (:)

```{r}
vecnum <- 1:10
vecnum
```

- Matrices are 2D objects
    - all values must be of the *same type*
    - you can create them with the `matrix()` function

```{r}
mat <- matrix(1:9)
mat
mat <- matrix(data=1:9,nrow=3,ncol=3)
mat
```

- Data Frames are also 2D objects
    - but columns can be of *different types*
    - you can create them with the `data.frame()` function

```{r}
df <- data.frame(1:3, c("a", "b", "c"))
df
# ugh, horrible default column names! let's change this:
df <- data.frame(col1=1:3, col2=c("a", "b", "c"))
df
```

# HELP!

One of the reasons why R became so popular is because there's lots
of useful functions (especially for statistics) and lots of ways to find help.
Every function out there has a help page. You can access those from
within R by typing "?" before the funcion name.

```{r}
?read.table
```


# What is a function?

A function is a chunk of code that takes in some performs a certain action.
It usually takes and **input** (also called **argument**), manipulates it, and spits out an **output**.

For example, the function *matrix* that takes in at least one argument, which is a vector of data, and creates a matrix with one column. Alternatively, you can give *matrix* two other numbers as input. The first one is the number of rows, and the second is the number of columns. The output is an object of type **matrix**.

> **Q1:** What does the function *sum()* take as input, an what does it return?

> **A:**

> **Tip:** You can guess by running the example below, and reading the help page.

```{r}
sum(1:10)
?sum
```

# How can we create a function?

The structure for creating a function is:
```{r}
name_of_function <- function(argument){ 
  #manipulate arguments
  result <- argument*10
  return(result)
}

#here name_of_function, argument and result can have any name
```

Writing a function is like writing a recipe: once you are done writing, it does not mean your food is ready. 

To actually have food, you need to read the recipe and perform all tasks.
Similarly, to actually get the result of your function, you need to **run** your function:

```{r}
name_of_function(argument=10)
```

- functions can have multiple arguments:
```{r}
myfunction2 <- function(arg1, arg2, arg3){ 
  #manipulate arguments
  r1 <- arg1*100
  r2 <- arg2*10
  r3 <- arg3*1
  myres <- c(r1,r2,r3)
  return(myres)
}
```

>**Q2:** How many arguments does the function above take?

>**A:**


>**Q3:** What type of object does it return?

>**A:**


