---
title: "Lab 9"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

# Natural Selection I

## Selection in infinite, haploid populations (p.130-132)

Assuming a haploid population, we can predict allele frequencies in a biallelic locus as a function of time, given the relative fitness of individuals with alleles *A* and *a*.

**Q1** Assume a single biallelic locus in a haploid population, with allele frequencies 0.01 for *A* and 0.99 for *a*, and relative fitness of 1 for *A* and 0.99 for *a*, as specified below. Calculate the expected frequency of allele *A* at the 100th generation after selection starts acting. What is the expected frequency of *A*?

**A: **

```{r}
# allele frequencies
fA <- 0.01
fa <- 1-fA
# relative fitness of alleles
wA <- 1
wa <- 0.99
# selection coefficient
s <- wA-wa

### Code your answer here
```

**Q2** Look at the *for* loop below and understand how it works to calculate the frequency of the *A* allele in generations 1 to 1000, given the same fitness and starting allele frequencies as in the previous exercise (given below). Check that you get approximately the same frequency of *A* at the 100th generation as you calculated above. What frequency did you get at the 100th generation?
We saw these equations for fA' you saw in lecture. See also page 131 of the book.

**A: **

```{r}
# allele frequencies
fA <- 0.01
fa <- 1-fA
# relative fitness of alleles
wA <- 1
wa <- 0.99
# selection coefficient
s <- 1-0.99

# Create a vector to store the allele frequencies in 1000 generations
ngen <- 1000
fA <- numeric(ngen)
fA[1] <- 0.01

for (gen in 2:ngen){
  fA[gen] <- ( wA * fA[gen-1] ) / (wA * fA[gen-1] + wa * (1-fA[gen-1]) )
}

# Look at position 100 in vector of allele frequencies fA to answer the question             

```


The code below should reproduce Figure 7.2 from the book
```{r}
# Reproducing Fig. 7.2 from the book:
plot(fA, 
     type = "l", # make this a line plot
     ylab = "fA(t)",
     xlab = "t (in generations)")
```

**Q3** Is the *A* allele deleterious or beneficial?

**A:** 


**Q4** What frequency does the *A* allele approach over time?

**A:** 

**Q5** Does the frequency of *A* change faster when *A* is at low (~0.01), intermediate (~0.5) or high (~0.99) frequency? Tip: Use the plot to answer this question.

**A:** 


## Selection in infinite, diploid populations (p.133-144)

In diploid organisms, each of the three possible genotypes in a biallelic locus can have different probabilities of survival into adulthood (i.e. viability), which translate into a different fitness for each genotype.

**Q6** Complete the for loop below to compute the frequency of allele *A* in the next 50 generations. Assuming the genotype fitness below and starting frequency of *A* of 0.01.
*Tip:* You will need to use one of the equations for fA' you saw in lecture. See also page 136 of the book.
```{r}
# fitness of each genotype
waa <- 0.6
wAa <- 0.8
wAA <- 1

ngen <- 50
fA <- numeric(ngen)
fA[1] <- 0.01

for (gen in 2:ngen){
# UNCOMMENT THE FOLLOWING LINE AND COMPLETE IT (add as many lines of code and intermediate variables as you see need for):
 # fA[gen] <-
}
```


If you calculated the frequencies correctly in the previous question, the code below should reproduce Figure 7.4 from the book.

```{r}
# Reproducing Fig. 7.2 from the book:
plot(fA, 
     type = "l",
     ylab = "fA(t)", 
     xlab = "t (in generations)")
```


**Q7** What type of selection is this? (Directional, overdominance or underdominance?) How is it generally defined in terms of the fitness of genotypes?

**A: ** 


**Q8** Which allele is advantageous? What frequency does this allele approach over time?

**A: **


### Selection on dominant vs. recessive alleles

**Q9** Use your for loop from **Q6** again to compute the frequency of *A* in the next 300 generations, in a scenario where *a is dominant* and *A* is recessive. Fitness of AA is 1 and fitness of aa is 0.9. Use initial frequency of *A* of 0.05. Plot the result.

```{r, eval=F}
# genotypic fitness
wAA <- 1
wAa <- # what is this if a is the dominant allele?
waa <- 0.9

# vector to store frequency of A in 300 generations
ngen <- 300
fA <- numeric(ngen)
# starting frequency of A in generation 1
fA[1] <- 0.05

# write your for loop below to compute fA in the next 299 generations

###########
# Plotting tricks used below:
# If you add `par(new=T)` before you plot fA, the lines are overlayed in the same plot. 
# If you add `par(mfrow=c(3,1))` it plots 3 plots in the same figure. 
# To specify colors to your line plot you can use `col='red'` as one of the plot arguments. 
# To add a legend, check up the help for the legend function: `?legend`

# plot your result
plot(fA, 
     type = "l",
     ylim = c(0,1),
     ylab = "fA(t)",
     xlab = "t (in generations)",
     col = "red")
legend(250,0.3,"A recessive",col="red",lty = 1)
```


**Q10** At around how many generations did *A* reach a frequency of 0.2 in the previous simulation? (You can estimate it from the plot, by eye.)

**A:** 


**Q11** Use your for loop from **Q6** again to compute the frequency of *A* through 300 generations, in a scenario where *A is dominant* and *a* is recessive. wAA is 1 and waa is 0.9. Use initial frequency of *A* of 0.05. Plot the result. 

```{r,eval=F}
wAA <- 1
wAa <-  # what is this if A is the dominant allele?
waa <- 0.9

# vector to store frequency of A in 300 generations
ngen <- 300
fA <- numeric(ngen)
# starting frequency of A in generation 1
fA[1] <- 0.05

# write your for loop below to compute fA in the next 299 generations

###########
# Plotting tricks used below:
# If you add `par(new=T)` before you plot fA, the lines are overlayed in the same plot. 
# If you add `par(mfrow=c(3,1))` it plots 3 plots in the same figure. 
# To specify colors to your line plot you can use `col='red'` as one of the plot arguments. 
# To add a legend, check up the help for the legend function: `?legend`

# plot your result
par(new=T)
plot(fA, 
     type = "l",
     ylim = c(0,1),
     ylab = "fA(t)",
     xlab = "t (in generations)",
     col = "blue")
legend(250,0.2,"A dominant",col="blue",lty = 1)
```

**Q12** At around how many generations did *A* reach a frequency of 0.2 in the previous simulation? (You can estimate it from the plot, by eye.)

**A:** 

**Q13** Comparing directional selection on recessive alleles to directional selection on dominant alleles, answer:

**Q13a** When an allele is at low frequency (0-0.2), is selection more efficient if the allele is dominant or recessive? In other words, in which scenario does selection change fA from 0.05 to 0.2 faster? Why?

**A:** 


**Q13b** When an allele is at high frequency (0.8-1), why does it take so long for selection to increase the frequency of a beneficial allele if it is dominant?

**A:** 


**Q14** Run the following code to simulate a population starting with frequency of *A* of 0.1 and keeping wAA=wAa=1. Then, vary waa from 0.1 to 0.5 to 0.8 and answer: How do different values of fitness of the disadvantageous genotype (aa, in this case) affect the rate of change of allele frequencies over time?

**A:** 

```{r}
wAA <- 1
wAa <- 1
waa <- 0.1
ngen <- 100
fA <- numeric(ngen)
fA[1] <- 0.1
for (gen in 2:ngen){
  v_bar <-  wAA*fA[gen-1]^2 + 2 * wAa * fA[gen-1] *(1-fA[gen-1]) + waa * (1-fA[gen-1])^2
  fA[gen] <- ( wAA * fA[gen-1]^2 + wAa * fA[gen-1]*(1-fA[gen-1]) ) / v_bar
}
plot(fA, 
     type = "l",
     ylab = "fA(t)",
     xlab = "t (in generations)",
     ylim=c(0,1),
     col="red")


waa <- 0.5
fA <- numeric(ngen)
fA[1] <- 0.1
for (gen in 2:ngen){
  v_bar <-  wAA*fA[gen-1]^2 + 2 * wAa * fA[gen-1] *(1-fA[gen-1]) + waa * (1-fA[gen-1])^2
  fA[gen] <- ( wAA * fA[gen-1]^2 + wAa * fA[gen-1]*(1-fA[gen-1]) ) / v_bar
}
par(new=T)
plot(fA, 
     type = "l",
     ylab = "fA(t)",
     xlab = "t (in generations)",
     ylim=c(0,1),
     col="blue")


waa <- 0.8
fA <- numeric(ngen)
fA[1] <- 0.1
for (gen in 2:ngen){
  v_bar <-  wAA*fA[gen-1]^2 + 2 * wAa * fA[gen-1] *(1-fA[gen-1]) + waa * (1-fA[gen-1])^2
  fA[gen] <- ( wAA * fA[gen-1]^2 + wAa * fA[gen-1]*(1-fA[gen-1]) ) / v_bar
}
par(new=T)
plot(fA, 
     type = "l",
     ylab = "fA(t)",
     xlab = "t (in generations)",
     ylim=c(0,1),
     col="green")

```


Run the following code to compute the frequency of *A* through 300 generations, with the following starting conditions and initial frequency of *A* of 0.01. Plot the result.

```{r}
wAA <- 0.8
wAa <- 1
waa <- 0.7

ngen <- 300
fA <- numeric(ngen)
fA[1] <- 0.01

for (gen in 2:ngen){
  v_bar <-  wAA*fA[gen-1]^2 + 2 * wAa * fA[gen-1] *(1-fA[gen-1]) + waa * (1-fA[gen-1])^2
  fA[gen] <- ( wAA * fA[gen-1]^2 + wAa * fA[gen-1]*(1-fA[gen-1]) ) / v_bar
}

plot(fA, 
     type = "l",
     ylab = "fA(t)",
     xlab = "t (in generations)")
```

**Q15** What frequency does allele *A* approach over time in this scenario? What about allele *a*?

**A:** 


**Q16** What type of selection is this? (Directional, overdominance or underdominance?) How is it generally defined in terms of the fitness of genotypes?

**A:** 

Run the following code to compute the frequency of *A* through 100 generations, with the following starting conditions and initial frequencies of *A* of 0.6 and 0.7. Plot the result.


```{r}
wAA <- 0.9
wAa <- 0.8
waa <- 1

ngen <- 100
fA <- numeric(ngen)


fA[1] <- 0.6

for (gen in 2:ngen){
  v_bar <-  wAA*fA[gen-1]^2 + 2 * wAa * fA[gen-1] *(1-fA[gen-1]) + waa * (1-fA[gen-1])^2
  fA[gen] <- ( wAA * fA[gen-1]^2 + wAa * fA[gen-1]*(1-fA[gen-1]) ) / v_bar
}

plot(fA, 
     type = "l",
     ylab = "fA(t)",
     xlab = "t (in generations)",ylim=c(0,1))

fA[1] <- 0.7

for (gen in 2:ngen){
  v_bar <-  wAA*fA[gen-1]^2 + 2 * wAa * fA[gen-1] *(1-fA[gen-1]) + waa * (1-fA[gen-1])^2
  fA[gen] <- ( wAA * fA[gen-1]^2 + wAa * fA[gen-1]*(1-fA[gen-1]) ) / v_bar
}

par(new=T)
plot(fA, 
     type = "l",
     ylab = "fA(t)",
     xlab = "t (in generations)",ylim=c(0,1))
par(new=F)
```

**Q17** Does the *A* allele get fixed or lost over time when initial allele frequency is 0.6? What if initial frequency of *A* is 0.7?

**A:** 

**Q18** What type of selection is this? (Directional, overdominance or underdominance?) How is it generally defined in terms of the fitness of genotypes?

**A:** 

